Query Optimization for Dynamic Imputation

### 2017-07-20 paper/examples.tex
paper: bold all figure labels
small fix: italicize Impute/Drop when referring to operator, for consistency

\begin{figure}
\begin{lstlisting}[language=SQL]
SELECT income, AVG(white_blood_cell_ct)
FROM demo, exams, labs
WHERE gender = 2 AND 
      weight >= 120 AND
      demo.id = exams.id AND 
      exams.id = labs.id
GROUP BY demo.income
\end{lstlisting}
\caption{A typical public health query on CDC's NHANES data.}
\label{fig:example-query}
\end{figure}

\begin{figure}
  \begin{subfigure}{\columnwidth}
    \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
  \caption{A quality-optimized plan for the query in~\Cref{fig:example-query}.}
\label{fig:quality-plan}
\end{subfigure}
  \begin{subfigure}{\columnwidth}
\vspace{1em}
\Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$}
        [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
            [.$\sigma_{\text{demo.gender} = 2}$ demo ]
            [.$\sigma_{\text{exams.weight} \geq 120}$ [.\colorbox{pink}{$\delta_{\text{exams.weight}}$} exams ] ] ] labs ] ] ] ]
\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.}
\label{fig:fast-plan}
\end{subfigure}
\vspace{1em}

@@

For the query in~\Cref{fig:example-query}, \ProjectName{} generates
several plans on an optimal frontier. \Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operator $\mu$,
which employs a reference imputation strategy to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, though other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
On the other hand, \Cref{fig:fast-plan} shows a performance-optimized plan that uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data in any field, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

>>>>

\begin{figure}
  \captionsetup{labelfont=bf}
\begin{lstlisting}[language=SQL]
SELECT income, AVG(white_blood_cell_ct)
FROM demo, exams, labs
WHERE gender = 2 AND 
      weight >= 120 AND
      demo.id = exams.id AND 
      exams.id = labs.id
GROUP BY demo.income
\end{lstlisting}
\caption{A typical public health query on CDC's NHANES data.}
\label{fig:example-query}
\end{figure}

\begin{figure}
  \captionsetup{labelfont=bf}
  \begin{subfigure}{\columnwidth}
    \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
  \caption{A quality-optimized plan for the query in~\Cref{fig:example-query}.}
\label{fig:quality-plan}
\end{subfigure}
  \begin{subfigure}{\columnwidth}
\vspace{1em}
\Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$}
        [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
            [.$\sigma_{\text{demo.gender} = 2}$ demo ]
            [.$\sigma_{\text{exams.weight} \geq 120}$ [.\colorbox{pink}{$\delta_{\text{exams.weight}}$} exams ] ] ] labs ] ] ] ]
\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.}
\label{fig:fast-plan}
\end{subfigure}
\vspace{1em}

@@

For the query in~\Cref{fig:example-query}, \ProjectName{} generates
several plans on an optimal frontier. \Cref{fig:quality-plan} shows a quality-optimized plan that uses the \textit{Impute} operator $\mu$,
which employs a reference imputation strategy to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, though other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
On the other hand, \Cref{fig:fast-plan} shows a performance-optimized plan that uses the \textit{Drop} operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data in any field, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

### 2017-07-20 paper/introduction.tex
small fix: italicize Impute/Drop when referring to operator, for consistency

These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The \textit{Impute} operation fills in missing data values using any statistical imputation
  technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
  The \textit{Drop} operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the
  \textit{quality} of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto-optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: \textit{Impute} ($\mu$) and \textit{Drop} ($\delta$).
  The \textit{Impute} operation fills in missing data values using any statistical imputation
  technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
  The \textit{Drop} operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the
  \textit{quality} of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto-optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}
